\vspace{30mm} {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\ {\large Document Version \DuoPLICRev}

### 2023-02-28 doc/src/IOMMU.tex

\textbf{%
Warning!
This chapter is only a draft, and might change significantly before
being ratified as standard by the {\RISCV} International Association.%
}
\bigskip

>>>>

\textbf{%
This chapter is \emph{frozen}, meaning
a functional change is extremely unlikely.
Change will occur only because of some truly critical issue
being identified during public review.%
}
\bigskip

### 2023-02-28 doc/src/VSLevel.tex
Clarify values supported by IID field of CSR hvictl
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

\textbf{%
This chapter is considered stable but could still change before being
ratified as standard by the {\RISCV} International Association.%
}
\bigskip

@@

All \z{hvictl} fields together can affect
the value of CSR \z{vstopi} (Virtual Supervisor Top Interrupt) and
therefore the interrupt identity reported in \z{vscause} when an
interrupt traps to \mbox{VS-mode}.
IID is a {\WARL} unsigned integer field with at least 6 implemented
bits, while IPRIO is always the full 8~bits.

>>>>

\textbf{%
This chapter is \emph{frozen} and has already passed public review,
making a functional change at this stage extremely unlikely.%
}
\bigskip

@@

All \z{hvictl} fields together can affect
the value of CSR \z{vstopi} (Virtual Supervisor Top Interrupt) and
therefore the interrupt identity reported in \z{vscause} when an
interrupt traps to \mbox{VS-mode}.
IID is a {\WARL} unsigned integer field with at least 6 implemented
bits, while IPRIO is always the full 8~bits.
If $k$ bits are implemented for IID, then all values 0 through
$\mbox{2}^{k}-\mbox{1}$ are supported, and a write to \z{hvictl}
sets IID equal to bits ($\mbox{15}+k$):16 of the value written.

### 2023-02-28 doc/src/intro.tex

Extensions Smaia and Ssaia cover only
those AIA features that impact the ISA at a hart.
Although the following are described or discussed
in this document as part of the AIA, they are not implied by
Smaia or Ssaia because the components are categorized as non-ISA:
APLICs, Duo-PLICs, IOMMUs, and any mechanisms for initiating
interprocessor interrupts apart from writing to IMSICs.

>>>>

Extensions Smaia and Ssaia cover only
those AIA features that impact the ISA at a hart.
Although the following are described or discussed
in this document as part of the AIA, they are not implied by
Smaia or Ssaia because the components are categorized as non-ISA:
APLICs, IOMMUs, and any mechanisms for initiating
interprocessor interrupts apart from writing to IMSICs.

### 2023-02-28 doc/src/preface.tex
Allow mvien + mvip virtual interrupts SEI, SSI
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

{
\begin{table}[hbt]
\centering
\begin{tabular}{|l|c|c|}
\hline
Chapter                                                  & ISA? & Status \\
\hline
\hline
1.\ Introduction                                         & ---  & Frozen \\
2.\ Control and Status Registers (CSRs) Added to Harts   & Yes  & Frozen \\
3.\ Incoming MSI Controller (IMSIC)                      & Yes  & Frozen \\
4.\ Advanced Platform-Level Interrupt Controller (APLIC) & No   & Frozen \\
5.\ Duo-PLIC                                             & No   & Draft \\
6.\ Interrupts for Machine and Supervisor Levels         & Yes  & Frozen \\
7.\ Interrupts for Virtual Machines (VS Level)           & Yes  & Frozen \\
8.\ Interprocessor Interrupts (IPIs)                     & No   & Frozen \\
9.\ IOMMU Support for MSIs to Virtual Machines           & No   & Draft \\
\hline
\end{tabular}
\end{table}
}

An implementation adhering to the current document might not conform
to an eventual ratified Advanced Interrupt Architecture for {\RISCV}.
The chapters marked in the table as \emph{Frozen} are not expected
to change significantly before being put up for ratification.
The chapters marked \emph{Draft} are very much
at risk of changing before ratification.

Concerning the two chapters that remain in draft state:
\begin{itemize}

\item
The authors are not aware that any implementation of
a Duo-PLIC yet exists, even one just for simulation.
So long as that remains true, the Duo-PLIC specification of
Chapter~\ref{ch:DuoPLIC} will not advance beyond draft status.

\item
A separate standard is being developed for IOMMUs in {\RISCV} systems.
Chapter~\ref{ch:IOMMU} about an IOMMU's support for MSIs
(message-signaled interrupts) is expected to be finalized
only in conjunction with the full {\RISCV} IOMMU specification.
\end{itemize}

\section*{Changes for RC2 (Ratification Candidate 2)}

>>>>

{
\begin{table}[hbt]
\centering
\begin{tabular}{|l|c|c|}
\hline
Chapter                                                  & ISA? & Status \\
\hline
\hline
1.\ Introduction                                         & ---  & Frozen \\
2.\ Control and Status Registers (CSRs) Added to Harts   & Yes  & Frozen \\
3.\ Incoming MSI Controller (IMSIC)                      & Yes  & Frozen \\
4.\ Advanced Platform-Level Interrupt Controller (APLIC) & No   & Frozen \\
5.\ Interrupts for Machine and Supervisor Levels         & Yes  & Frozen \\
6.\ Interrupts for Virtual Machines (VS Level)           & Yes  & Frozen \\
7.\ Interprocessor Interrupts (IPIs)                     & No   & Frozen \\
8.\ IOMMU Support for MSIs to Virtual Machines           & No   & Frozen \\
\hline
\end{tabular}
\end{table}
}

An implementation adhering to the current document might not conform
to an eventual ratified Advanced Interrupt Architecture for {\RISCV}.
However, all chapters are \emph{frozen}, meaning they are not expected
to change significantly before being put up for ratification.

\section*{Changes for RC3 (Ratification Candidate 3)}

Removed the still-draft Duo-PLIC chapter to a separate document.

In Section~\ref{sec:virtIntrs-S} added the options
for bits 1 and 9 to be writable in CSR \z{mvien},
and specified the effects of setting each of these bits.

Upgraded Chapter~\ref{ch:IOMMU} (``IOMMU Support'')
to the \emph{frozen} state.

\section*{Changes for RC2}

### 2023-02-28 doc/src/riscv-duoplic.tex

%=======================================================================
\chapter{Duo-PLIC}
\label{ch:DuoPLIC}

\textbf{%
Warning!
This chapter is only a draft, and might change significantly before
being accepted as standard by the {\RISCV} International Association.
It remains possible that the Duo-PLIC specification
will never become a ratified standard.%
}
\bigskip

The Advanced PLIC of the previous chapter is not backward compatible
with the original {\RISCV} PLIC.
To ease the transition of software from the original PLIC to the
APLIC, a {\RISCV} system may have a \emph{\mbox{Duo-PLIC\/}}
that software can configure to act as either form of PLIC.
A \mbox{Duo-PLIC} is an extension of an APLIC, having all the
usual memory-mapped interfaces of an APLIC, plus, at some other
memory addresses, the memory-mapped interface of an original {\RISCV}
PLIC too.
When the \mbox{Duo-PLIC} has \emph{compatibility mode} enabled,
interrupts from the \mbox{Duo-PLIC} to harts are controlled exclusively
by the original PLIC side of the \mbox{Duo-PLIC};
and when compatibility mode is disabled, interrupts to harts
are controlled exclusively by the APLIC side of the
\mbox{Duo-PLIC}.
A \mbox{Duo-PLIC} comes out of reset with compatibility mode enabled,
so older software with no knowledge of an APLIC can see the
\mbox{Duo-PLIC} as an original PLIC.

@@

\item
Interrupts to harts are determined by the APLIC side of the
\mbox{Duo-PLIC}, as specified in Chapter~\ref{ch:AdvPLIC}.
The state of the original PLIC side of the \mbox{Duo-PLIC} is
irrelevant.

@@

However, an original PLIC encodes priorities in the opposite order
from an IMSIC or APLIC, with number 1 being interpreted as the
lowest priority and larger numbers as higher priorities.
When a \mbox{Duo-PLIC} is acting as an original PLIC and reports to
a hart an interrupt priority number (see Chapter~\ref{ch:MSLevel}
concerning a hart's handling of interrupt priorities), the number
reported is negated to \emph{normalize} it to the convention of the
Advanced Interrupt Architecture.
If the original PLIC's priority number for an interrupt is~$p$,
the normalized priority number communicated to the target hart is
${\mbox{2}^\textrm{IPRIOLEN}-p}$.

\begin{commentary}
Negation of original PLIC priority numbers to normalize them matters
only when a hart supports configurability of the priorities of major
interrupts, an option added by the Advanced Interrupt Architecture.
See Sections \ref{sec:intrPrios-M} and~\ref{sec:intrPrios-S}.
Software that is oblivious to the Advanced Interrupt Architecture
should have no reason to see priority numbers being negated.
\end{commentary}

>>>>

%=======================================================================
% riscv-interrupts.tex
%-----------------------------------------------------------------------

\documentclass[twoside,11pt]{article}

% Fix copy/pasting of ligatures in Acrobat
\input{glyphtounicode.tex}
\pdfgentounicode=1 %

\input{preamble}

\newcommand{\DuoPLICRev}{0.3.3-draft}

\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}

%=======================================================================

\begin{document}

%=======================================================================

\title{%
  \vspace{30mm}
  {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\
  {\large Document Version \DuoPLICRev}
}

\author{%
  Editor: John Hauser \\
  {\tt jh.riscv\mbox{}@jhauser.us} \\
  \\
  \today
}

\date{}
\maketitle

\thispagestyle{empty}

\vfill

\textbf{%
Warning!
This document is only a draft and might never be
accepted by the {\RISCV} International Association.
If accepted, the Duo-PLIC specification here is
liable to change before becoming a ratified standard.%
}

\vfill

Contributors to all versions of this specification in alphabetical
order (please contact the editor to suggest corrections):
John Hauser.

This document is released under a Creative Commons Attribution 4.0
International License.

\vfill

\clearpage

%=======================================================================

The Advanced PLIC of the {\RISCV} Advanced Interrupt Architecture
is not backward compatible with the original {\RISCV} PLIC.
To ease the transition of software from the original PLIC to the
APLIC, a {\RISCV} system may have a \emph{\mbox{Duo-PLIC\/}}
that software can configure to act as either form of PLIC.
A \mbox{Duo-PLIC} is an extension of an APLIC, having all the
usual memory-mapped interfaces of an APLIC, plus, at some other
memory addresses, the memory-mapped interface of an original {\RISCV}
PLIC too.
When the \mbox{Duo-PLIC} has \emph{compatibility mode} enabled,
interrupts from the \mbox{Duo-PLIC} to harts are controlled exclusively
by the original PLIC side of the \mbox{Duo-PLIC};
and when compatibility mode is disabled, interrupts to harts
are controlled exclusively by the APLIC side of the
\mbox{Duo-PLIC}.
A \mbox{Duo-PLIC} comes out of reset with compatibility mode enabled,
so older software with no knowledge of an APLIC can see the
\mbox{Duo-PLIC} as an original PLIC.

@@

\item
Interrupts to harts are determined by
the APLIC side of the \mbox{Duo-PLIC}.
The state of the original PLIC side of the \mbox{Duo-PLIC} is
irrelevant.

@@

However, an original PLIC encodes priorities in the opposite order
from an IMSIC or APLIC, with number 1 being interpreted as the
lowest priority and larger numbers as higher priorities.
When a \mbox{Duo-PLIC} is acting as an original PLIC
and reports to a hart an interrupt priority number,
the number reported is negated to \emph{normalize} it to
the convention of the Advanced Interrupt Architecture.
If the original PLIC's priority number for an interrupt is~$p$,
the normalized priority number communicated to the target hart is
${\mbox{2}^\textrm{IPRIOLEN}-p}$.

\begin{commentary}
Negation of original PLIC priority numbers to normalize them matters
only when a hart supports configurability of the priorities of major
interrupts, an option added by the Advanced Interrupt Architecture.
Software that is oblivious to the Advanced Interrupt Architecture
should have no reason to see priority numbers being negated.
\end{commentary}

%=======================================================================

\end{document}
